---
title: "Getting Started with SeqPlotR"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Getting Started with SeqPlotR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 3.5,
  dev        = "png",
  dpi        = 192
)
set.seed(1)
```

```{css echo = FALSE}
img { max-width: 100%; height: auto; }
```

SeqPlotR builds genomic figures by composing a plot, tracks, and
elements with a single `%+%` operator — modelled on ggplot2 but
specialised for `GRanges` data. This vignette walks through the
minimum you need to render a plot end-to-end.

```{r load}
library(SeqPlotR)
library(GenomicRanges)
```

## The four primitives you need

A SeqPlotR figure is always built from these pieces:

| Piece | What it does |
|-------|-------------|
| `seq_plot()` | Creates a new plot. |
| `seq_track()` | Adds a genomic panel with a `windows` range. |
| `map(...)` | Declares data-driven aesthetics (`x`, `y`, `color`, ...). |
| `aes(...)` | Declares constant aesthetics and theme keys. |

Elements (`seq_point`, `seq_line`, `seq_area`, ...) are drawn inside a
track. The `%+%` operator chains everything together and dispatches on
the right-hand side's class.

## A first plot

```{r first}
win <- GRanges("chr1", IRanges(1, 1e6))
gr  <- GRanges("chr1",
               IRanges(start = seq(1e4, 9.9e5, length.out = 200), width = 1),
               score = sin(seq(0, 6*pi, length.out = 200)) +
                       rnorm(200, 0, 0.1))

p <- seq_plot() %+%
  seq_track(data = gr, mapping = map(x = start, y = score), windows = win) %+%
  seq_point(aesthetics = aes(color = "#205EA6", size = 0.4))

p$plot()
```

The track holds the defaults (data, mapping, windows); the element
`seq_point()` inherits them.

## Layering several elements in one track

`%+%` on a plot adds elements to the **most recently added track**, so
two elements in a row draw on top of each other:

```{r layers}
gr_line <- GRanges("chr1",
                   IRanges(seq(1e4, 9.9e5, length.out = 200), width = 1),
                   value = cumsum(rnorm(200)) * 0.05)

p <- seq_plot() %+%
  seq_track(data = gr_line, mapping = map(x = start, y = value), windows = win) %+%
  seq_area(aesthetics = aes(fill = "#92BFDB", color = "#205EA6", baseline = 0)) %+%
  seq_point(aesthetics = aes(color = "#205EA6", size = 0.3))

p$plot()
```

## Two tracks side by side: `%|%` and `%__%`

`%|%` and `%__%` are convenience aliases for `%+%` that set
`direction = "right"` or `direction = "under"` on the track they
introduce:

```{r sidebyside, fig.height = 5}
gr_a <- GRanges("chr1", IRanges(seq(1e4, 9.9e5, length.out = 100), width = 1),
                score = rnorm(100))
gr_b <- GRanges("chr1", IRanges(seq(1e4, 9.9e5, length.out = 100), width = 1),
                score = rexp(100, rate = 0.5))

p <- seq_plot() %|%
  seq_track(data = gr_a, mapping = map(x = start, y = score),
            windows = win, track_id = "A") %+% seq_line(aesthetics = aes(color = "#205EA6")) %__%
  seq_track(data = gr_b, mapping = map(x = start, y = score),
            windows = win, track_id = "B") %+% seq_area(aesthetics = aes(fill = "#F9AE77"))

p$plot()
```

## Printing

`seq_plot` carries an S3 `print` method, so a bare expression in a
script or knitr chunk auto-renders — `p$plot()` is equivalent.

## Where to go next

- `vignette("patchwork-layouts", "SeqPlotR")` — multi-cell layouts
  with a layout string.
- `vignette("wrappers", "SeqPlotR")` — higher-level wrappers
  (`seq_copynumber`, `seq_hic`, `seq_chip`, ...).
- `vignette("links", "SeqPlotR")` — cross-track arcs, arches,
  synteny, and zoom links.
