---
title: "Legends in SeqPlotR"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Legends in SeqPlotR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 3.2,
  dev       = "png",
  dpi       = 192
)
set.seed(1)
```

```{css echo = FALSE}
img { max-width: 100%; height: auto; }
```

SeqPlotR attaches legend metadata directly to the elements that draw the data.
A `LegendKey` describes one glyph-plus-label row; a `SeqLegendSpec` says where
and how to lay those rows out. For continuous color scales, `seq_gradient_legend()`
produces a `GradientLegendSpec` that renders as a filled color bar with a tick
axis. Three position families are supported:

| Position | Where it renders |
|----------|-----------------|
| `"inside"` | Overlaid on the data panel — no margin space required. |
| `"track_margin"` | In the canvas margin alongside a specific track. |
| `"canvas_margin"` | Aggregated from all tracks into one shared margin band. |

---

## Data used in this vignette

```{r load}
library(SeqPlotR)
library(GenomicRanges)

# 1-Mb window; axis labels will read 0.0 – 1.0 (scaled by 1e-6)
win <- GRanges("chr1", IRanges(1, 1e6))

# Synthetic ChIP-seq signal: sparse background with 8 sharp peaks
make_chip <- function(n = 200, n_peaks = 8, seed = 1) {
  set.seed(seed)
  starts <- sort(sample(1:999000, n))
  scores <- runif(n, 0, 0.15)                        # flat baseline
  peaks  <- sort(sample(seq_len(n), n_peaks))
  for (pk in peaks) {
    w <- max(1, pk - 3):min(n, pk + 3)
    scores[w] <- scores[w] + runif(length(w), 0.4, 0.85)
  }
  GRanges("chr1", IRanges(starts, width = 1), score = pmin(scores, 1))
}

gr_a <- make_chip(seed = 1)   # H3K27ac-like
gr_b <- make_chip(seed = 2)   # H3K4me3-like

# Continuous-score data for gradient legend examples
set.seed(42)
n_pts  <- 120
gr_cnt <- GRanges(
  "chr1",
  IRanges(sort(sample(1:999000, n_pts)), width = 1),
  score = round(runif(n_pts, 0, 100), 1)
)
```

---

## 1. Inside legends

Inside legends are overlaid directly on the data panel. They require no
extra margin space.

### Quick placement with a bare `LegendKey`

Pass a bare `LegendKey` to `legend` and it is automatically wrapped into an
inside legend centered in the panel (`x = 0.5, y = 0.5`). No `seq_legend()`
call is needed:

```{r inside-bare, fig.cap = "A bare LegendKey is auto-placed at the panel centre."}
p1 <- seq_plot() %+%
  seq_track(data = gr_a, mapping = map(x = start, y = score),
            windows = win) %+%
  seq_area(
    aesthetics = aes(fill = "#D08080", alpha = 0.8),
    legend     = LegendKey(label = "H3K27ac",
                           color = "#A02020", fill = "#D08080", shape = "-")
  )
p1$plot()
```

### Controlling placement with `seq_legend()`

Use `seq_legend()` to pin the legend to a specific corner. `x` and `y` are
fractions of the panel width and height; `y = 0.95` places the anchor near
the top:

```{r inside-anchored, fig.cap = "seq_legend() anchors the legend to the top-left corner with a title."}
k1 <- LegendKey(label = "H3K27ac", color = "#A02020", fill = "#D08080", shape = "-")
k2 <- LegendKey(label = "H3K4me3", color = "#205080", fill = "#8090C0", shape = "-")

spec_inside <- seq_legend(
  list(k1, k2),
  title       = "ChIP marks",
  position    = "inside",
  x           = 0.02,   # 2% from the left edge
  y           = 0.95,   # 95% of the way up the panel
  orientation = "vertical"
)

p2 <- seq_plot() %+%
  seq_track(data = gr_a, mapping = map(x = start, y = score),
            windows = win) %+%
  seq_area(
    aesthetics = aes(fill = "#D08080", alpha = 0.8),
    legend     = spec_inside
  )
p2$plot()
```

### Right-aligned horizontal layout

`hjust = 1` right-aligns the key block; `orientation = "horizontal"` puts all
keys in one row:

```{r inside-horizontal, fig.cap = "Two keys in a horizontal row, right-aligned at the top."}
spec_h <- seq_legend(
  list(k1, k2),
  title       = "ChIP marks",
  position    = "inside",
  x           = 0.98, hjust = 1,   # right-align
  y           = 0.95,
  orientation = "horizontal"
)

p3 <- seq_plot() %+%
  seq_track(data = gr_a, mapping = map(x = start, y = score),
            windows = win) %+%
  seq_area(
    aesthetics = aes(fill = "#D08080", alpha = 0.8),
    legend     = spec_h
  )
p3$plot()
```

---

## 2. Track-margin legends

`"track_margin"` draws the legend in the canvas margin *beside* one
specific track. Set `side` to choose the edge and enlarge the matching
plot margin so there is room to render:

```{r track-margin, fig.cap = "Two-key vertical legend in the 18% left canvas margin beside the track."}
spec_tm <- seq_legend(
  list(k1, k2),
  title       = "ChIP marks",
  position    = "track_margin",
  side        = "left",
  orientation = "vertical",
  x           = 0.1,    # fraction of margin width
  y           = 0.8     # fraction of margin height (0 = bottom, 1 = top)
)

p_tm <- seq_plot(
  aesthetics = aes(margins = list(top   = 0.03, bottom = 0.03,
                                   left  = 0.18, right  = 0.03))
) %+%
  seq_track(data = gr_a, mapping = map(x = start, y = score),
            windows = win) %+%
  seq_area(
    aesthetics = aes(fill = "#D08080", alpha = 0.8),
    legend     = spec_tm
  )
p_tm$plot()
```

`orientation` defaults to `"vertical"` for `side %in% c("left", "right")` and
`"horizontal"` for top / bottom:

```{r margin-orientation}
seq_legend(k1, position = "track_margin", side = "left")$orientation
seq_legend(k1, position = "track_margin", side = "top")$orientation
```

---

## 3. Canvas-margin legends (aggregated)

`"canvas_margin"` collects legend entries from **all tracks** and renders them
once, spanning the full width of the chosen margin side. Keys from specs that
share the same `title` are merged under one heading.

Use `%__%` instead of `%+%` to stack the second track below the first
(`direction = "under"`). Without stacking both tracks would sit side-by-side
in a single row:

```{r canvas-margin, fig.height = 5, fig.cap = "Two stacked tracks sharing a single canvas-margin legend at the top."}
# Each spec carries the key for one track.
# The first spec's x / hjust controls alignment of the aggregated block.
spec_ca <- seq_legend(k1, title = "ChIP marks", position = "canvas_margin",
                      side = "top", x = 0.5, hjust = 0.5)
spec_cb <- seq_legend(k2, title = "ChIP marks", position = "canvas_margin",
                      side = "top")

p_cm <- seq_plot(
  aesthetics = aes(margins = list(top   = 0.12, bottom = 0.04,
                                   left  = 0.04, right  = 0.04))
) %+%
  seq_track(data = gr_a, mapping = map(x = start, y = score),
            windows = win) %+%
  seq_area(aesthetics = aes(fill = "#D08080", alpha = 0.8), legend = spec_ca) %__%
  seq_track(data = gr_b, mapping = map(x = start, y = score),
            windows = win) %+%
  seq_area(aesthetics = aes(fill = "#8090C0", alpha = 0.8), legend = spec_cb)

p_cm$plot()
```

When track titles differ, `canvas_margin` inserts a separator between groups.
Here the two tracks use different titles to show the grouping:

```{r canvas-margin-groups, fig.height = 5, fig.cap = "Two title groups separated in the canvas margin."}
spec_ca2 <- seq_legend(k1, title = "Active enhancers",
                       position = "canvas_margin", side = "top",
                       x = 0.5, hjust = 0.5)
spec_cb2 <- seq_legend(k2, title = "Promoters",
                       position = "canvas_margin", side = "top")

p_cm2 <- seq_plot(
  aesthetics = aes(margins = list(top   = 0.14, bottom = 0.04,
                                   left  = 0.04, right  = 0.04))
) %+%
  seq_track(data = gr_a, mapping = map(x = start, y = score),
            windows = win) %+%
  seq_area(aesthetics = aes(fill = "#D08080", alpha = 0.8), legend = spec_ca2) %__%
  seq_track(data = gr_b, mapping = map(x = start, y = score),
            windows = win) %+%
  seq_area(aesthetics = aes(fill = "#8090C0", alpha = 0.8), legend = spec_cb2)

p_cm2$plot()
```

---

## 4. Gradient color bars

Continuous color scales need a color bar rather than discrete key rows.
`seq_gradient_legend()` produces a `GradientLegendSpec` that renders as a
filled rectangle graded from the palette's low to high color, with a tick-and-
label axis derived from the data limits.

The placement arguments (`position`, `x`, `y`, `hjust`, `orientation`, `side`)
work identically to `seq_legend()`.

### Manual color bar inside the panel

```{r gradient-inside, fig.cap = "Vertical viridis color bar pinned to the right side of the panel."}
grad_inside <- seq_gradient_legend(
  palette     = "viridis",
  limits      = c(0, 100),
  title       = "Score",
  position    = "inside",
  x           = 0.97,          # near the right edge
  y           = 0.5,           # vertically centred
  hjust       = 1,
  orientation = "vertical"
)

p_gi <- seq_plot() %+%
  seq_track(data = gr_cnt, mapping = map(x = start, y = score),
            windows = win) %+%
  seq_point(
    aesthetics = aes(color = "#2166AC", size = 1.2),
    legend     = grad_inside
  )
p_gi$plot()
```

### Color bar in the track margin

Place the bar in the right track margin for a clean separation from the data:

```{r gradient-track-margin, fig.cap = "Horizontal plasma color bar in the right track margin."}
grad_tm <- seq_gradient_legend(
  palette     = "plasma",
  limits      = c(0, 100),
  title       = "Score",
  position    = "track_margin",
  side        = "right",
  orientation = "vertical",
  x           = 0.5,
  y           = 0.5
)

p_gtm <- seq_plot(
  aesthetics = aes(margins = list(top    = 0.03, bottom = 0.03,
                                   left   = 0.03, right  = 0.20))
) %+%
  seq_track(data = gr_cnt, mapping = map(x = start, y = score),
            windows = win) %+%
  seq_point(
    aesthetics = aes(color = "#2166AC", size = 1.2),
    legend     = grad_tm
  )
p_gtm$plot()
```

### Auto-legend from `map()`

When a continuous column is mapped via `map()` on a supported element
(`SeqBar`, `SeqLine`, `SeqArea`, `SeqPoint`, `SeqSegment`, `SeqTile`,
`SeqRect`), a `GradientLegendSpec` is generated automatically — no
`seq_gradient_legend()` call is needed. The bar defaults to
`position = "track_margin"`, `side = "right"`.

Pass `show_legend = FALSE` to opt out for a specific element:

```{r gradient-auto, fig.cap = "Auto color bar from map(fill = score); no seq_gradient_legend() needed."}
p_auto <- seq_plot(
  aesthetics = aes(margins = list(top    = 0.03, bottom = 0.03,
                                   left   = 0.03, right  = 0.20))
) %+%
  seq_track(data = gr_cnt, mapping = map(x = start, y = score),
            windows = win) %+%
  seq_bar(
    mapping    = map(fill = score),
    aesthetics = aes(alpha = 0.9)
    # show_legend = FALSE   # uncomment to suppress the auto bar
  )
p_auto$plot()
```

### Controlling tick breaks

`breaks` accepts three forms:

- `NULL` (default) — `pretty()` with five ticks.
- A single integer — `pretty()` targeting that many ticks.
- A numeric vector — exact break values.

```{r gradient-breaks, fig.cap = "Color bar with explicit breaks at 0, 25, 50, 75, 100."}
grad_breaks <- seq_gradient_legend(
  palette     = "magma",
  limits      = c(0, 100),
  breaks      = c(0, 25, 50, 75, 100),
  title       = "Score",
  position    = "inside",
  x           = 0.97,
  y           = 0.5,
  hjust       = 1,
  orientation = "vertical"
)

p_brk <- seq_plot() %+%
  seq_track(data = gr_cnt, mapping = map(x = start, y = score),
            windows = win) %+%
  seq_point(
    aesthetics = aes(color = "#8B3A8B", size = 1.2),
    legend     = grad_breaks
  )
p_brk$plot()
```

---

## 5. Suppression

Legends can be silenced at three granularities. Each level overrides all
finer-grained settings below it.

### Element level

`show_legend = FALSE` on one element silences only that element, leaving
others in the same track unaffected:

```{r suppress-element}
el_shown  <- seq_area(legend = LegendKey(label = "Foreground"))
el_hidden <- seq_area(legend = LegendKey(label = "Background"),
                      show_legend = FALSE)

is.null(el_shown$collect_legend_keys())   # FALSE — contributes
is.null(el_hidden$collect_legend_keys())  # TRUE  — suppressed
```

This also suppresses auto-generated gradient bars:

```{r suppress-element-auto}
el_no_bar <- seq_bar(mapping = map(fill = score), show_legend = FALSE)
is.null(el_no_bar$auto_legend)   # TRUE — no bar generated
```

### Track level

`show_legend = FALSE` on a `seq_track()` silences every element in that
track, regardless of individual element settings:

```{r suppress-track}
trk <- seq_track(windows = win,
                 elements  = list(el_shown, el_hidden),
                 show_legend = FALSE)
is.null(trk$collect_legend_keys())   # TRUE — whole track suppressed
```

### Plot level

Pass `show_legend = FALSE` (or the shorthand `legend = FALSE`) to
`seq_plot()` to suppress all legend output for the entire plot, including
auto-generated gradient bars:

```{r suppress-plot, fig.cap = "legend = FALSE on seq_plot() removes all legend output."}
p_no_legend <- seq_plot(legend = FALSE) %+%
  seq_track(data = gr_a, mapping = map(x = start, y = score),
            windows = win) %+%
  seq_area(
    aesthetics = aes(fill = "#D08080", alpha = 0.8),
    legend     = spec_inside   # spec defined earlier — ignored here
  )
p_no_legend$plot()
```

---

## 6. Inspecting legend metadata: `collect_legend_keys()`

`collect_legend_keys()` aggregates legend entries from an element, a track,
or (transitively) all tracks in a plot. It returns `NULL` when nothing
contributes, or a list of named lists:

| Field | Content |
|-------|---------|
| `title` | Group heading (from `LegendKey$title` or `SeqLegendSpec$title`). |
| `key` | The `LegendKey` object. |
| `element_class` | R6 class name of the contributing element. |

```{r collect-element}
k  <- LegendKey(label = "ChIP signal", title = "Tracks", color = "darkgreen")
el <- seq_area(legend = k)
str(el$collect_legend_keys()[[1]])
```

Track-level aggregation flattens all element entries:

```{r collect-track}
el1 <- seq_area(legend = LegendKey(label = "H3K27ac", color = "#A02020"))
el2 <- seq_area(legend = LegendKey(label = "H3K4me3", color = "#205080"))

trk  <- seq_track(windows = win, elements = list(el1, el2))
keys <- trk$collect_legend_keys()
cat(length(keys), "keys from track:\n")
for (entry in keys)
  cat(" ", entry$key$label, "—", entry$element_class, "\n")
```

---

## Summary

| Task | How |
|------|-----|
| Attach a legend, auto-placed at panel centre | `seq_area(legend = LegendKey(...))` |
| Anchor to a specific corner | `seq_legend(keys, position="inside", x=, y=)` |
| Draw beside one track | `seq_legend(keys, position="track_margin", side=)` + enlarge that margin |
| Aggregate from all tracks into one band | `seq_legend(keys, position="canvas_margin", side=)` |
| Add a continuous color bar (manual) | `seq_gradient_legend(palette=, limits=, ...)` |
| Add a continuous color bar (auto) | `map(fill = score)` or `map(color = score)` on any supported element |
| Suppress one element | `seq_area(show_legend = FALSE)` |
| Suppress a whole track | `seq_track(show_legend = FALSE)` |
| Suppress the entire plot | `seq_plot(legend = FALSE)` or `seq_plot(show_legend = FALSE)` |
| Inspect what will be shown | `element$collect_legend_keys()` / `track$collect_legend_keys()` |

| `position` | Best for |
|-----------|---------|
| `"inside"` | Single-track plots; compact per-track annotation |
| `"track_margin"` | When you want to keep the data area clean on one track |
| `"canvas_margin"` | Multi-track plots with a shared legend group |

### Gradient legend quick reference

| Argument | Effect |
|----------|--------|
| `palette` | `"viridis"`, `"plasma"`, `"magma"`, `"blues"`, `"reds"` |
| `limits` | Data range the gradient spans, e.g. `c(0, 100)` |
| `breaks = NULL` | Auto ticks via `pretty()` (default) |
| `breaks = 5L` | Request 5 ticks via `pretty()` |
| `breaks = c(0,50,100)` | Exact tick positions |
| `orientation = "vertical"` | Tall bar with right-side axis |
| `orientation = "horizontal"` | Wide bar with bottom axis |
