---
title: "New features: expansion, OOB, titles, margins, inheritance"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{New features: expansion, OOB, titles, margins, inheritance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 3.2,
  dev        = "png",
  dpi        = 192
)
options(seqplotr.suppress_oob = FALSE)
set.seed(7)
```

```{css echo = FALSE}
img { max-width: 100%; height: auto; }
```

```{r load, message = FALSE}
library(SeqPlotR)
library(GenomicRanges)
```

This vignette covers the cross-cutting feature batch that touches scales,
axes, margins, and the Hi-C fetcher:

1. Axis **expansion** now actually applies to element coordinates.
2. New `oob = "exclude"` / `"perimeter"` argument on every scale.
3. `aes(axis.x.line = NA)` removes a piece of axis chrome, ggplot-style.
4. Axis titles auto-derive from the **first element's** mapping.
5. Genomic axes default each window's title to its **seqname**.
6. Manual title overrides via `axis.<side>.title.label` /
   `axis.<side>.title = aes(label = …, col = …)`.
7. New `pretty = list(...)` argument forwards to base `pretty()`.
8. `seq_plot(windows = …)` is inherited by tracks; genomic scales
   inherit windows from their enclosing track.
9. New `seq_plot(plot_margin = 0.02)` outermost canvas margin.
10. New `seq_plot(highlight_margins = TRUE)` development overlay.
11. `open_hic()$fetch()` now accepts a multi-range `region`; the
    `max_fetch_bp` default raised to 280 Mb.

We use a synthetic two-track example throughout. The data:

```{r data}
win <- GRanges("chr1", IRanges(start = 1, end = 1e6))
n   <- 200
gr  <- GRanges("chr1", IRanges(start = sample(1:1e6, n), width = 1))
mcols(gr)$xv    <- runif(n, 0, 10)
mcols(gr)$score <- runif(n, 0, 6)
mcols(gr)$grp   <- sample(letters[1:3], n, TRUE)
```

# 1. Axis expansion (now wired through)

Previously, passing `expand = c(0.1, 0)` on a continuous scale had no
visible effect — the breaks were computed but the npc transform that
positions data on the panel used the raw data range. As of this batch,
`expand` is honoured end-to-end: the panel carries an `xplot_range` /
`yplot_range` that includes the expansion, and elements map into it.

```{r expand}
trk_noexp <- seq_track(
  data    = gr, windows = win,
  mapping = map(x = xv, y = score),
  scale_x = seq_scale_continuous(expand = c(0,   0)),
  aesthetics = aes("axis.x.title.label" = "x (no expand)")
) %+% seq_point()

trk_exp <- seq_track(
  data    = gr, windows = win,
  mapping = map(x = xv, y = score),
  scale_x = seq_scale_continuous(expand = c(0.2, 0)),
  aesthetics = aes("axis.x.title.label" = "x (20% expansion)")
) %+% seq_point()

(seq_plot() %+% trk_noexp %+% trk_exp)$plot()
```

# 2. `oob` — out-of-bounds handling

The new `oob` argument controls what happens when data falls outside the
(expanded) axis range. `"exclude"` (default) drops the rows and reports
a count via `message()`; `"perimeter"` clamps them to the limit and
reports.

```{r oob}
# A subset whose x runs from -3 to 13 — three points sit outside [0, 10].
gr_off <- gr
mcols(gr_off)$xv <- jitter(c(-3, mcols(gr)$xv[2:(n-1)], 13), amount = 1)

trk_excl <- seq_track(
  data    = gr_off, windows = win,
  mapping = map(x = xv, y = score),
  scale_x = seq_scale_continuous(limits = c(0, 10), oob = "exclude"),
  aesthetics = aes("axis.x.title.label" = "oob = exclude")
) %+% seq_point()

trk_per <- seq_track(
  data    = gr_off, windows = win,
  mapping = map(x = xv, y = score),
  scale_x = seq_scale_continuous(limits = c(0, 10), oob = "perimeter"),
  aesthetics = aes("axis.x.title.label" = "oob = perimeter")
) %+% seq_point()

(seq_plot() %+% trk_excl %+% trk_per)$plot()
```

# 3. `NA` removes axis chrome

Setting any of the structural axis sub-keys (`line`, `title`, `ticks`,
`labels`, `text`, `gridline`) to `NA` blanks that piece — the same
contract as ggplot's `element_blank()`. Sides `x` / `y` (default both),
`x1` / `x2` / `y1` / `y2` (specific axis), and the top-level
`axis.*` (every axis) all work.

```{r na-blank}
trk_full <- seq_track(
  data = gr, windows = win,
  mapping = map(x = xv, y = score)
) %+% seq_point()

trk_no_line_no_ticks <- seq_track(
  data = gr, windows = win,
  mapping = map(x = xv, y = score),
  aesthetics = aes(
    "axis.x.line"  = NA,
    "axis.x.ticks" = NA,
    "axis.x.text"  = NA
  )
) %+% seq_point()

(seq_plot() %+% trk_full %+% trk_no_line_no_ticks)$plot()
```

# 4 & 5 & 6. Smarter axis titles

The default axis title now derives from the first element's mapping
(falling back to the track's mapping, then to the seqname when the scale
is genomic). You can override with a plain string label or a styled
`aes(label = ..., col = ..., size = ...)`.

```{r titles-default}
# Element mapping wins over track mapping for the title.
trk_a <- seq_track(data = gr, windows = win) %+%
         seq_point(mapping = map(x = xv, y = score))

# Manual override via the convenience .label sub-key.
trk_b <- seq_track(data = gr, windows = win,
                   mapping    = map(x = xv, y = score),
                   aesthetics = aes("axis.x.title.label" = "Custom title")) %+%
         seq_point()

# Or a full styled aes — sets label and any nested style keys at once.
trk_c <- seq_track(
  data       = gr, windows = win,
  mapping    = map(x = xv, y = score),
  aesthetics = aes(axis.x.title = aes(label = "Styled title",
                                       col   = "#A0282C",
                                       size  = 1.0))
) %+% seq_point()

(seq_plot() %+% trk_a %+% trk_b %+% trk_c)$plot()
```

Genomic axes (`seq_scale_genomic()`, or the auto-built genomic scale
when the user supplies no `scale_x`) get a per-window seqname as the
default title:

```{r titles-genomic}
multi_win <- GRanges(
  c("chr1", "chr2", "chr3"),
  IRanges(start = c(1, 1, 1), end = c(1e6, 1e6, 1e6))
)
multi_data <- GRanges(
  rep(c("chr1", "chr2", "chr3"), each = 50),
  IRanges(start = c(sample(1:1e6, 50),
                    sample(1:1e6, 50),
                    sample(1:1e6, 50)), width = 1)
)
mcols(multi_data)$score <- runif(150, 0, 5)

trk_multi <- seq_track(
  data    = multi_data, windows = multi_win,
  mapping = map(y = score)
) %+% seq_point()

(seq_plot() %+% trk_multi)$plot()
```

# 7. `pretty = list(...)`

All position scales now accept a `pretty` argument: a named list of
additional arguments forwarded to base `pretty()`.

```{r pretty}
trk_default <- seq_track(
  data = gr, windows = win,
  mapping    = map(x = xv, y = score),
  scale_x    = seq_scale_continuous(limits = c(0, 10)),
  aesthetics = aes("axis.x.title.label" = "default n_breaks = 5")
) %+% seq_point()

trk_pretty <- seq_track(
  data = gr, windows = win,
  mapping    = map(x = xv, y = score),
  scale_x    = seq_scale_continuous(
                  limits = c(0, 10),
                  pretty = list(n = 10, min.n = 5)),
  aesthetics = aes("axis.x.title.label" = "pretty = list(n = 10, min.n = 5)")
) %+% seq_point()

(seq_plot() %+% trk_default %+% trk_pretty)$plot()
```

# 8. Windows inheritance

`seq_plot()` accepts a `windows` argument that any track without its own
windows inherits. Genomic scales constructed without explicit `windows`
inherit from the enclosing track. This trims boilerplate when many
tracks share the same windows.

```{r windows-inherit}
# Old style — every track repeats the windows.
old <- seq_plot() %+%
  (seq_track(data = gr, windows = win,
             mapping = map(x = xv, y = score)) %+% seq_point()) %+%
  (seq_track(data = gr, windows = win,
             mapping = map(x = xv, y = score)) %+% seq_point())

# New style — windows live on the plot.
new <- seq_plot(windows = win) %+%
  (seq_track(data = gr,
             mapping = map(x = xv, y = score)) %+% seq_point()) %+%
  (seq_track(data = gr,
             mapping = map(x = xv, y = score)) %+% seq_point())

new$plot()
```

Genomic scales also pick up the enclosing track's windows when
constructed bare:

```{r scale-inherit}
trk_bare_scale <- seq_track(
  data    = gr, windows = win,
  scale_x = seq_scale_genomic()           # no windows arg
) %+% seq_point()

(seq_plot() %+% trk_bare_scale)$plot()
```

# 9. `plot_margin`

`seq_plot(plot_margin = 0.02)` sets the outermost canvas margin (in NPC
units). The default is `0.02`. Per-side overrides via
`aesthetics = aes(margins = list(top = ..., ...))` still win.

```{r plot-margin}
tight  <- seq_plot(plot_margin = 0,    windows = win) %+%
          (seq_track(data = gr,
                     mapping = map(x = xv, y = score)) %+% seq_point())
roomy  <- seq_plot(plot_margin = 0.08, windows = win) %+%
          (seq_track(data = gr,
                     mapping = map(x = xv, y = score)) %+% seq_point())

tight$plot()
roomy$plot()
```

# 10. `highlight_margins` — development overlay

A development aid: setting `highlight_margins = TRUE` overlays
translucent rectangles on every margin band. Useful when tuning the
layout or chasing a stray gap.

```{r highlight-margins}
highlighted <- seq_plot(windows = win, highlight_margins = TRUE, plot_margin = 0.05) %+%
  (seq_track(data = gr,
             mapping = map(x = xv, y = score),
             track_outer_margin  = 0.04,
             track_inner_margin  = 0.03,
             window_outer_margin = 0.02,
             window_inner_margin = 0.02) %+%
   seq_point())
highlighted$plot()
```

Colours:

| Region                  | Colour       |
|-------------------------|--------------|
| `plot_margin`           | red          |
| `track_outer_margin`    | dark blue    |
| `track_inner_margin`    | light blue   |
| `window_outer_margin`   | dark green   |
| `window_inner_margin`   | light green  |
| plotting area (`inner`) | pink         |

All drawn at `alpha = 0.5`.

# 11. Hi-C fetcher: multi-range and 280 Mb default

`open_hic()$fetch()` now takes a single `region` GRanges that may
contain multiple ranges; each range produces its own intra-region
contact submatrix, and the results are concatenated into a single BEDPE
data frame. The previous `region1` / `region2` interface is replaced.

The `max_fetch_bp` default has been raised from 5 Mb to **280 Mb** so
whole-chromosome fetches work out of the box; the check now applies per
range rather than to the sum.

```r
hic <- open_hic("path/to/file.hic", resolution = 25000L)

# Single range — same as before, but uses the new arg name.
hic$fetch(GRanges("chr14", IRanges(98e6, 99e6)))

# Multi-range fetch in one call.
multi_regions <- GRanges(
  c("chr14", "chr15"),
  IRanges(start = c(98e6, 50e6), end = c(99e6, 51e6))
)
hic$fetch(multi_regions)
```

Each row in the returned data frame carries the seqname of the range
that produced it (`seqnames1`/`seqnames2`), so downstream code can still
distinguish per-region contacts.

# Putting it together

A small showcase that uses several of the new features in one plot:

```{r showcase, fig.height = 4.2}
showcase <- seq_plot(windows = multi_win, plot_margin = 0.03) %+%
  (seq_track(
     data    = multi_data,
     mapping = map(y = score),
     aesthetics = aes(axis.x.title = aes(label = "Position",
                                          col   = "#3D4D58"))
   ) %+% seq_point()) %+%
  (seq_track(
     data    = multi_data,
     mapping = map(x = start, y = score),
     scale_y = seq_scale_continuous(limits = c(0, 4),
                                     expand = c(0.05, 0),
                                     oob    = "exclude",
                                     pretty = list(n = 4))
   ) %+% seq_path())
showcase$plot()
```
