---
title: "Patchwork Layouts in SeqPlotR"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Patchwork Layouts in SeqPlotR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 5,
  dev        = "png",
  dpi        = 192
)
set.seed(1)
```

```{css echo = FALSE}
img { max-width: 100%; height: auto; }
```

SeqPlotR supports two layout modes:

- **Positional** — tracks flow using `direction = "right"` or
  `"under"` (aliased by the `%|%` and `%__%` operators).
- **Patchwork** — a multiline string assigns each track to a named
  rectangular region. This is inspired by the
  [`patchwork`](https://patchwork.data-imaginist.com) layout string.

```{r load}
library(SeqPlotR)
library(GenomicRanges)

win <- GRanges("chr1", IRanges(1, 1e6))
```

## Positional layout via track widths and heights

```{r positional}
gr <- GRanges("chr1", IRanges(seq(1, 1e6, length.out = 200), width = 1),
              score = rnorm(200))

p <- seq_plot() %+%
  seq_track(data = gr, mapping = map(x = start, y = score),
            windows = win, track_id = "A", track_width = 2) %+% seq_line() %|%
  seq_track(data = gr, mapping = map(x = start, y = score),
            windows = win, track_id = "B") %+% seq_point() %__%
  seq_track(data = gr, mapping = map(x = start, y = score),
            windows = win, track_id = "C") %+% seq_area()

p$plot()
```

`track_width` and `track_height` are *relative* within the row and
column respectively: track `A` above is twice as wide as `B`.

## Patchwork layout string

Passing a layout string to `seq_plot()` makes every cell position
explicit. Each unique letter is a region; `#` cells stay blank.

```{r patchwork-string}
layout <- "
##AA
##AA
BBBC
BBBD
"
p <- seq_plot(layout = layout) %+%
  seq_track(data = gr, mapping = map(x = start, y = score),
            windows = win, track_id = "A") %+% seq_line() %+%
  seq_track(data = gr, mapping = map(x = start, y = score),
            windows = win, track_id = "B") %+% seq_area()  %+%
  seq_track(data = gr, mapping = map(x = start, y = score),
            windows = win, track_id = "C") %+% seq_bar()   %+%
  seq_track(data = gr, mapping = map(x = start, y = score),
            windows = win, track_id = "D") %+% seq_point()

p$plot()
```

With a layout string, each track's `direction` is ignored — placement
is decided entirely by the `track_id` matching the letters.

## Previewing a layout before adding data

`seq_preview_layout()` is the fastest way to validate a layout string:

```{r preview}
seq_preview_layout(layout = layout)
```

You can also preview a positional plot built from the operator chain:

```{r preview-positional}
seq_preview_layout(plot_obj = p)
```

## Circular preview

`seq_preview_circos()` draws a minimal circular layout, useful as a
schematic for whole-genome plots. It extracts the track structure from
a positional `seq_plot` (patchwork layouts are not supported):

```{r circos, fig.width = 5, fig.height = 5}
gw  <- default_genome_windows()
circ <- seq_plot() %+%
  seq_track(windows = gw, track_id = "whole_genome") %+% seq_line()
seq_preview_circos(plot_obj = circ)
```

## Relative widths and heights revisited

Both positional and patchwork layouts respect `track_width` and
`track_height` *within* a cell; the outer cell bounds come from the
layout string.

```{r weights, fig.height = 6}
layout <- "
AAAB
AAAB
CCCC
"
p <- seq_plot(layout = layout) %+%
  seq_track(data = gr, mapping = map(x = start, y = score),
            windows = win, track_id = "A", track_height = 2) %+% seq_line() %+%
  seq_track(data = gr, mapping = map(x = start, y = score),
            windows = win, track_id = "B") %+% seq_bar() %+%
  seq_track(data = gr, mapping = map(x = start, y = score),
            windows = win, track_id = "C") %+% seq_point()

p$plot()
```
