---
title: "High-Level Wrappers: Copy Number, Hi-C, and ChIP"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{High-Level Wrappers: Copy Number, Hi-C, and ChIP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4,
  dev        = "png",
  dpi        = 192
)
set.seed(42)
```

```{css echo = FALSE}
img { max-width: 100%; height: auto; }
```

SeqPlotR ships a small family of wrapper functions that assemble
common genomic visualisations from the primitive elements. Each
wrapper returns a `seq_plot` that can be further composed with
`%+%`, `%|%`, `%__%`, or [`seq_resolve()`].

```{r load}
library(SeqPlotR)
library(GenomicRanges)
```

## `seq_copynumber()`

Single-sample CN scatter coloured by integer state, with optional
segmentation overlay.

```{r cn}
win <- GRanges(c("chr1", "chr2"), IRanges(c(1, 1), c(1e6, 1.5e6)))

cn_gr <- GRanges(
  rep(c("chr1", "chr2"), each = 150),
  IRanges(start = c(sample(1:1e6, 150), sample(1:1.5e6, 150)), width = 5000),
  cn        = sample(0:5, 300, replace = TRUE,
                     prob = c(.05, .1, .55, .15, .1, .05)),
  log2ratio = rnorm(300, 0, 0.3)
)

seg <- GRanges(c("chr1", "chr2"),
               IRanges(c(1, 1), c(1e6, 1.5e6)),
               seg_mean = c(0.0, -0.3))

p <- seq_copynumber(cn_gr, windows = win,
                    segment_data = seg, segment_col = "seg_mean")
p$plot()
```

Column detection is automatic — `cn_col` falls back through
`c("cn", "copy_number", "CN", "state", "integer_cn")` and `ratio_col`
through `c("log2ratio", "logR", "log2R", "ratio", "log2_ratio")`.

## `seq_cn_heatmap()`

Multi-sample CN heatmap — one row per sample, tiles coloured by CN
state.

```{r cn-heatmap, fig.height = 3.5}
samples <- paste0("S", 1:5)
cnh_gr  <- GRanges(
  rep(c("chr1", "chr2"), each = 300),
  IRanges(start = c(sample(1:1e6, 300, replace = TRUE),
                    sample(1:1.5e6, 300, replace = TRUE)), width = 5000),
  sample = sample(samples, 600, replace = TRUE),
  cn     = sample(0:5, 600, replace = TRUE)
)

p <- seq_cn_heatmap(cnh_gr, windows = win,
                    sample_col = "sample", cn_col = "cn",
                    sample_order = samples)
p$plot()
```

## `seq_hic()` — four styles

One call to `seq_hic()` produces one style. Combine styles via
`seq_resolve()` for side-by-side panels.

```{r hic-data, include = FALSE}
# Hi-C contact simulator ported from THEfunc
# (inst/examples/08_rotated_hic_heatmap.R): every (i, j) bin pair gets a
# contact strength = exp(-distance * decay_rate) * lognormal noise, then
# the symmetric (j, i) pair is mirrored in.
generate_hic_matrix <- function(n_bins, decay_rate = 0.2) {
  ij    <- expand.grid(i = seq_len(n_bins), j = seq_len(n_bins))
  upper <- ij$j >= ij$i
  ij    <- ij[upper, , drop = FALSE]
  d     <- ij$j - ij$i
  strength <- pmax(0.01,
                   exp(-d * decay_rate) * rlnorm(nrow(ij), 0, 0.3))
  upper_df <- data.frame(bin_i = ij$i, bin_j = ij$j, strength = strength)
  off      <- d > 0
  lower_df <- data.frame(bin_i = ij$j[off], bin_j = ij$i[off],
                         strength = strength[off])
  rbind(upper_df, lower_df)
}

# 50 bins x 100 kb = 5 Mb region, blocks aligned to a clean grid.
set.seed(123)
n_bins     <- 50
bin_size   <- 1e5
chrom      <- "chr5"
region_st  <- 40e6
bin_starts <- region_st + (seq_len(n_bins) - 1L) * bin_size

mat <- generate_hic_matrix(n_bins, decay_rate = 0.25)

hic_gr <- GRanges(chrom,
                  IRanges(start = bin_starts[mat$bin_i], width = bin_size),
                  i_start = bin_starts[mat$bin_i],
                  i_end   = bin_starts[mat$bin_i] + bin_size,
                  j_start = bin_starts[mat$bin_j],
                  j_end   = bin_starts[mat$bin_j] + bin_size,
                  score   = mat$strength)
win1   <- GRanges(chrom, IRanges(region_st, region_st + n_bins * bin_size))
```

### `style = "triangle"`

```{r hic-tri, fig.height = 3}
seq_hic(hic_gr, windows = win1, style = "triangle")$plot()
```

### `style = "rectangle"`

Caps the distance axis at `max_dist`:

```{r hic-rect, fig.height = 3}
seq_hic(hic_gr, windows = win1, style = "rectangle", max_dist = 3e5)$plot()
```

### `style = "full"` and `"diagonal"`

The full contact matrix lives on a 2D genomic grid:

```{r hic-full, fig.width = 5, fig.height = 5}
seq_hic(hic_gr, windows = win1, style = "full")$plot()
```

## `seq_chip()`

ChIP-style signal + peak tracks. `data` is either a named list of
`GRanges` (one per sample) or a single `GRanges` with a sample column.

```{r chip, fig.height = 5}
make_sig <- function(nm) {
  g <- GRanges("chr1",
               IRanges(sort(sample(1:1e6, 400)), width = 500),
               score = rexp(400, rate = 0.15))
  S4Vectors::mcols(g)$sample <- nm
  g
}
sig <- list(Rad21 = make_sig("Rad21"), NippedB = make_sig("NippedB"))
peaks <- list(Rad21   = GRanges("chr1", IRanges(sort(sample(1:1e6, 20)), width = 2000)),
              NippedB = GRanges("chr1", IRanges(sort(sample(1:1e6, 15)), width = 2000)))

p <- seq_chip(sig, peaks = peaks,
              windows = GRanges("chr1", IRanges(1, 1e6)))
p$plot()
```

## Composing wrappers with `seq_resolve()`

`seq_resolve()` is the key plumbing function for combining wrapper
plots — it extracts each child's tracks and appends them to a parent
`seq_plot` under the requested direction.

```{r resolve, fig.height = 6}
win1 <- GRanges("chr1", IRanges(1, 1e6))

cn1 <- seq_copynumber(cn_gr[seqnames(cn_gr) == "chr1"],
                      windows = win1, track_id = "CN1")
hic <- seq_hic(hic_gr, windows = win1, style = "triangle",
               track_id = "HiC")

fig <- seq_resolve(seq_plot(), cn1, hic)
fig$plot()
```

Pass unique `track_id` values to each wrapper call to avoid ID
collisions when resolving multiple plots into one figure.
